Finiteness in Cubical Type Theory

### 2020-06-30 sections/finiteness-predicates.tex
finished manifest bishop section

Going from manifest Bishop finiteness is relatively straightforward:
to construct a proof of split enumerability from one of manifest Bishop
finiteness, it suffices to convert a proof of \(x \inunique \mathit{xs}\) to
one of \(x \in \mathit{xs}\), for all \(x\) and \(\mathit{xs}\).
Since \(\inunique\) is defined as a contraction of \(\in\), such a conversion
is simply the \(\text{fst}\) function.

@@

We have now proved that every manifestly Bishop finite type is split enumerable,
and vice versa.
While the types are not \emph{equivalent} (there are more split enumerable
proofs than there are manifest Bishop finite proofs), they are of equal power.
\subsection{From Manifest Bishop Finiteness to Equivalence}

We have seen that split enumerability was in fact a split-surjection in
disguise.
We will now see that manifest Bishop finiteness is in fact an \emph{equivalence}
in disguise.
\begin{multicols}{2}
  \begin{agdalisting} \label{is-equiv-def}
    \ExecuteMetaData[agda/Snippets/Equivalence.tex]{is-equiv-def}
  \end{agdalisting} \columnbreak
  \begin{agdalisting} \label{equiv-def}
    \ExecuteMetaData[agda/Snippets/Equivalence.tex]{equiv-def}
  \end{agdalisting}
\end{multicols}

\begin{definition}[Equivalences] \label{equivalences}
  We will take contractible maps \cite[definition 4.4.1]{hottbook} as our
  ``default'' definition of equivalences.
\end{definition}
\begin{lemma} \label{bishop-equiv}
  Manifest bishop finiteness is equivalent to an equivalence to a finite prefix
  of the natural numbers.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{bishop-is-equiv-type}
  \end{agdalisting}
\end{lemma}

\begin{proof} \let\qed\relax \vspace{-1\baselineskip}
  \begin{minipage}[t]{.7\textwidth}\vspace{-\baselineskip}
    \begin{agdalisting*}
      \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{bishop-is-equiv}
    \end{agdalisting*}
  \end{minipage}
  \begin{minipage}[t]{.19\textwidth} \setstretch{1.1}

@@

% \begin{proof} \todo{Convert to Agda}
%   \begin{align*}
%      \AgdaDatatype{\ensuremath{\mathcal{B}}}(A) &
%     \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , x \inunique \mathit{xs}
%     && \text{def.~\ref{bish-def} }(\AgdaDatatype{\ensuremath{\mathcal{B}}})
%     \\
%     & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(x \in \mathit{xs})
%     && \text{eqn.~\ref{uniq-memb-def} } (\inunique)
%     \\
%     & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(\text{fib}_{\text{snd}(\mathit{xs})}(x))
%     && \text{eqn.~\ref{container-membership} } (\in)
%     \\
%     & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
%     && \text{eqn.~\ref{is-equiv-def} (isEquiv)}
%     \\
%     & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \AgdaDatatype{Fin} \rrbracket (A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
%     && \text{def.~\ref{list-def} } (\AgdaDatatype{List})
%     \\
%     & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \AgdaDatatype{Fin}\;n) , A) , \text{isEquiv}(\text{snd}(\mathit{xs}))
%     && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
%     \\
%     & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \AgdaDatatype{Fin}\;n \rightarrow A) , \text{isEquiv}(f)
%     && \text{Reassociation of } \Sigma
%     \\
%     & \simeq \Sigma (n : \mathbb{N}) , ( \AgdaDatatype{Fin}\;n \simeq A )
%     && \text{eqn.~\ref{equiv-def} } (\simeq) \; \qedhere
%   \end{align*}
% \end{proof}
This proof is almost identical\footnotemark to the proof for
lemma~\ref{split-enum-is-split-surj}: it reveals that
enumeration-based finiteness predicates are simply another perspective on
relation-based ones.
\footnotetext{
  Unfortunately in our formalisation this proof cannot be a single line: for
  performance reasons \(\simeq\) is defined as a record type with eta-equality
  disabled, instead of the definition here which uses \(\Sigma\).
}

>>>>

Going from manifest Bishop finiteness is relatively straightforward:
to construct a proof of split enumerability from one of manifest Bishop
finiteness, it suffices to convert a proof of \(x\;\AgdaFunction{\ensuremath{\in!}}\;\mathit{xs}\) to
one of \(x\;\AgdaFunction{\ensuremath{\in}}\;\mathit{xs}\), for all \(x\) and \(\mathit{xs}\).
Since \AgdaFunction{\(\in!\)} is defined as a contraction of \AgdaFunction{\(\in\)}, such a conversion
is simply the \AgdaField{fst} function.

@@

We have now proved that every manifestly Bishop finite type is split enumerable,
and vice versa.
While the types are not \emph{equivalent} (there are more split enumerable
proofs than there are manifest Bishop finite proofs), they are of equal power.
\subsection{From Manifest Bishop Finiteness to Equivalence}
We have seen that split enumerability was in fact a split-surjection in
disguise.
We will now see that manifest Bishop finiteness is in fact an \emph{equivalence}
in disguise.
We define equivalences as contractible maps \cite[definition 4.4.1]{hottbook}:
\begin{multicols}{2}
  \begin{agdalisting} \label{is-equiv-def}
    \ExecuteMetaData[agda/Snippets/Equivalence.tex]{is-equiv-def}
  \end{agdalisting} \columnbreak
  \begin{agdalisting} \label{equiv-def}
    \ExecuteMetaData[agda/Snippets/Equivalence.tex]{equiv-def}
  \end{agdalisting}
\end{multicols}
\begin{lemma} \label{bishop-equiv}
  Manifest bishop finiteness is equivalent to an equivalence to a finite prefix
  of the natural numbers.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{bishop-is-equiv-type}
  \end{agdalisting}
\end{lemma}
\begin{proof} \let\qed\relax \vspace{-1\baselineskip}
  \begin{minipage}[t]{.7\textwidth}\vspace{-\baselineskip}
    \begin{agdalisting*}
      \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{bishop-is-equiv}
    \end{agdalisting*}
  \end{minipage}
  \begin{minipage}[t]{.19\textwidth} \setstretch{1.1}

@@

This proof is almost identical\footnotemark to the proof for
lemma~\ref{split-enum-is-split-surj}: it reveals that
enumeration-based finiteness predicates are simply another perspective on
relation-based ones.
\footnotetext{
  Unfortunately in our formalisation this proof cannot be a single line: for
  performance reasons \(\simeq\) is defined as a record type with eta-equality
  disabled, instead of the definition here which uses \(\Sigma\).
}
